Manuscript title here

### 2024-09-29 project-plan.Rmd
la till metod

## Epidemiology
Trauma is a major global public health concern. It causes over four milion death and affects hundreds of millions each year, among those younger populations are the most affected (2). Globally low and middle income countries are disproportionately affected, but high income countries like Sweden also face considerable challenges(källa). In Sweden between 2018 and 2021, the most common age demographic for trauma is the working aged population, between 18 and 64 years old. (Swetrau ÅR, 2021). This not only leads to premature loss of life but also imposes a substantial socio-economic burden, due to long-term disabilities and rehabilitation needs.

## Trauma
Trauma are generally divided into two groups, blunt and penetrating injury. The most common type in Sweden is blunt injury, constituting around 90% of all trauma. The most common causes for blunt trauma are traffic related and falls, which account for roughly ½ and ⅓ of all blunt trauma(Swetrau ÅR, 2021).

@@

## Shock
The definition for shock is circulatory failure leading to insufficient perfusion and oxygenation of tissue and organ(9). Some of the common symptoms are tachycardia, hypotenstion and altered mental status. Shock are usually classified into four different types based on physiological mechanism. The types are, hypovolemic, obstructive, cardiogenic and distributive, and they all have their own special characteristics(källa). These causes can occur alone or in combination, as there can be multiple disruptions on the circulatory system. Common examples of diagnoses for each type are, hypovolemic - bleeding, obstructive - pulmonary embolism, cardiogenic - STEMI  and, distributive - sepsis(källa).

@@

## Shock classifications
To differentiate the severity of trauma shock, we can divide it into different degrees. One of such classification is the ATLS classification for major haemorrhage. ATLS stratifies patients into four classes, based on the patients: estimated blood loss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output. However, this only meant to be used for hypovolemic shocks and not the other types. There is other classifications for predicting massive transfusion (TASH,TBSS, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2). The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock. ATLS in this regard, even though having its own shortcomings(10), still is one of the most used classifications for this topic(Källa).

## Trauma quality improvement and oppertunities for improvement
The initial management of trauma patients once they reach hospital is very time sensitive and error prone (3) [(4)](5), and shock is the most common potentially preventable death within the first 24 hour of injury(6), as the midian time to death is 2 hours(Tisherman, 2015). To address these challenges and reduce preventable errors, different types of Quality Improvement (QI) initiatives were initiated. QI focuses on systematically evaluating healthcare processes and patient outcomes, aiming to reduce morbidity and mortality(WHO). By implementing formal quality measures and conducting collaborative case reviews, QI programs have strengthened trauma systems and improved patient outcomes in various settings globally (Juillard + Hashmi).

@@

If you are conducting an observational study, for example a cohort or
case control study in which you assess associations between some
exposure and an outcome then use
[STROBE](https://www.equator-network.org/reporting-guidelines/strobe/).*

@@

<!-- Present key elements of study design -->

We conducted a registry based retrospective cohort study, using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna. We then linked the two databases to assess what types of OFI occurred in patients arriving in shock and use logistic regression to assess the association between the degree of shock and odds of opportunities for improvement.

@@

The trauma registry includes about 14000 patients treated at Karolinska University Hospital in Solna between 2014 and 2023. The "problem" database is a subset of the trauma registry and includes about ? patients selected for review. The Karolinska University Hospital in Solna treats all major trauma in the greater metropolitan area of Stockholm.

@@

To qualify into the trauma registry, requires admission by trauma team activation or any admission with an Injury Severity Score (ISS) of more than nine.

Every trauma patient at Karolinska University Hospital is included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters. Patients identified with a high potential for OFIs are discussed at multidisciplinary conferences. The identified OFIs are then categorized into broader areas. The multidisciplinary conferences are held every six to eight weeks, during which an average of ten patient cases are reviewed by experienced specialists from all trauma-related fields. The presence or absence of OFIs is determined by consensus among all participants and is documented in the trauma care quality database.

We included all patients in "SWETRAU" and "Problem" between 2014 and 2023 We excluded patients younger than 15 and/or were dead on arrival.

@@

<!-- Elaborate on the outcome OFI. How is it defined? What are the different types of OFI? -->

The outcome was defined as presence of atleast one OFI, determined by the multidisciplinary M&M conference in the "problem" registry. An OFI can be various types of errors, which are categorized into: clinical judgement error, inadequate resources, delay in treatment, missed injury, inadequate protocols, "preventable death" (den här var väl inte ens med i "problem", fast fanns med i studien som Martin delade på discord) and other errors.

In this study, the outcomes will be studied in two ways, one is the amount of occurrence for various types of OFI in shock patients, and the other is binary where odds of OFI for every shock class i calculated out of presence of OFI or not. (kanske inte ens en del av metod)

@@

<!-- Elaborate on how you will categorise the degree of shock, what are the different levels?-->

Due to lack of data on FAST, type of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[@yucel_trauma_2006][@ogura_modified_2016][@hanlin_prehospital_2024].

Version 1, 
Due to the tenth edition of ATLS classifications lack actual numbers, we will be using simplified and numerical version approximated by Dunham et al[@dunham_comparison_2017]. (INSERT PICTURE)

Version 2, 
Due to the tenth edition of ATLS classifications lack actual numbers, we will be approximating numbers to the values. For HR and RR we will be using the numbers from the ninth edition of the classification. The SBP will be divded into class I and II where the SBP is >110, class III 110-90mmhg and class IV <90mmhg. This selection is based on tenth edition, where SBP is defined as normal i class I to II and normal/below in class III and clearly below normal i class IV. These two studies, redefined hypotension as 110mmhg respective 90mmhg dependent on age[@eastridge_hypotension_2007][@oyetunji_redefining_2011], therefore we choose class III as 110-90mmgh and class IV below 90mmhg. For mental status, according to the tenth edition ATLS, class I and II should have normal GCS, and class III and IV have lower GCS. Therefore we decided to use GCS 15 for class I + II and GCS <15 for class III and IV.

 - Hur gör vi med de som hamnar mellan klasserna?
To classify, all of the parameters in one class needs to be fulfilled. In the cases where the parameters belong in different classes, we will be classifying according to the most severe parameter in the same way as Dunham et al[@dunham_comparison_2017]. At same time ATLS manual states one of the common pitfalls is "Diagnosis of shock can be missed when only a single parameter is used." therefore to improve sensitivity we decided to use the worst parameter for classification.

To prevent classifying non-shock traumatic brain injury as class III and worse shock based on GCS, we decided to only include those who also present a circulatory parameter in accordance with class III or IV. 
 - eller ta bort GCS helt och hållet? För i enlighet med ovan kommer GCS alltid att prioriteras bort och då finns ingen anlending att ha med det...

Version 3, simplaste och den jag gillar mest, men kan andra sjukdomar som gluykolen förgiftning sabba tanken?
We will be only using the base deficit parameter according to the tenth edition of the ATLS classification for haemorrhage. There are three reasons. The first one is ATLS tenth edition lack numerical values for vital parameters.  The second one is that bace deficit is superior to vital parameters as a predictive parameter for mortality[@dunham_comparison_2017]. The third one is due to nature of having multiple parameters, makes it difficult to categorise a patient with vital parameters in two or more classes and difficult to separate lower GCS due to traumatic brain injury from shock as cause. The classes will be divided into class I: 0 - (-2), class II: (-2)-(-6), class III: (-6)-(-10), and class IV: <(-10).

@@

@@

All available data in "SWETRAU" and "Problem" database will be included.

@@

The statistical analysis will be performed using R, a programing language and environment for statistical computing. We will present the type of OFI as percentage distributions and then visualize it in a pie chart. Bivariable logistic regression will be used to determine the association between the OFI and degree/class of shock according to ATLS. It will be presented as odds ratios (OR) between presence of OFI, and shock classes. The OR will be determined with 95% confidence intervals, and a significance level of 5% will be used. All statistical analysis will first be done on synthetic data and later implemented on the data collected from the trauma registry and the “problem” database to ensure objectivity.

@@

*This is part C) in the project plan [sv: projektredogörelse]*

>>>>

## Epidemiology

Trauma is a major global public health concern. It causes over four milion death and affects hundreds of millions each year, among those younger populations are the most affected (2). Globally low and middle income countries are disproportionately affected, but high income countries like Sweden also face considerable challenges(källa). In Sweden between 2018 and 2021, the most common age demographic for trauma is the working aged population, between 18 and 64 years old. (Swetrau ÅR, 2021). This not only leads to premature loss of life but also imposes a substantial socio-economic burden, due to long-term disabilities and rehabilitation needs.

## Trauma

Trauma are generally divided into two groups, blunt and penetrating injury. The most common type in Sweden is blunt injury, constituting around 90% of all trauma. The most common causes for blunt trauma are traffic related and falls, which account for roughly ½ and ⅓ of all blunt trauma(Swetrau ÅR, 2021).

@@

## Shock

The definition for shock is circulatory failure leading to insufficient perfusion and oxygenation of tissue and organ(9). Some of the common symptoms are tachycardia, hypotenstion and altered mental status. Shock are usually classified into four different types based on physiological mechanism. The types are, hypovolemic, obstructive, cardiogenic and distributive, and they all have their own special characteristics(källa). These causes can occur alone or in combination, as there can be multiple disruptions on the circulatory system. Common examples of diagnoses for each type are, hypovolemic - bleeding, obstructive - pulmonary embolism, cardiogenic - STEMI  and, distributive - sepsis(källa).

@@

## Shock classifications

To differentiate the severity of trauma shock, we can divide it into different degrees. One of such classification is the ATLS classification for major haemorrhage. ATLS stratifies patients into four classes, based on the patients: estimated blood loss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output. However, this only meant to be used for hypovolemic shocks and not the other types. There is other classifications for predicting massive transfusion (TASH,TBSS, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2). The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock. ATLS in this regard, even though having its own shortcomings(10), still is one of the most used classifications for this topic(Källa).

## Trauma quality improvement and oppertunities for improvement

The initial management of trauma patients once they reach hospital is very time sensitive and error prone (3) [(4)](5), and shock is the most common potentially preventable death within the first 24 hour of injury(6), as the midian time to death is 2 hours(Tisherman, 2015). To address these challenges and reduce preventable errors, different types of Quality Improvement (QI) initiatives were initiated. QI focuses on systematically evaluating healthcare processes and patient outcomes, aiming to reduce morbidity and mortality(WHO). By implementing formal quality measures and conducting collaborative case reviews, QI programs have strengthened trauma systems and improved patient outcomes in various settings globally (Juillard + Hashmi).

@@

@@

We conducted a registry based retrospective cohort study, using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.

@@

The trauma registry includes patients treated at Karolinska University Hospital in Solna between 2014 and 2023. The trauma care quality database is a subset of the trauma registry and includes patients selected for review. The Karolinska University Hospital in Solna treats all major trauma in the greater metropolitan area of Stockholm.

@@

We included all patients in trauma registry and Trauma care quality database. We excluded patients younger than 15 and/or were dead on arrival.

The inclusion criteria for the trauma registry is a patient either admitted through trauma team activation or a already admitted patient with an Injury Severity Score (ISS) greater than nine. Every trauma patient (registered in the trauma registry) at Karolinska University Hospital is then included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters. Patients identified with a high potential for OFIs are discussed at multidisciplinary conferences. The identified OFIs are then categorized into broader areas. The multidisciplinary conferences are held every six to eight weeks, during which an average of ten patient cases are reviewed by experienced specialists from all trauma-related fields. The presence or absence of OFIs is determined by consensus among all participants and is documented in the trauma care quality database.

@@

The outcome was defined as the presence of at least one OFI, determined by the multidisciplinary M&M conference in the “problem” registry. An OFI can be various types of errors, which are categorized into: clinical judgement error, inadequate resources, delay in treatment, missed injury, inadequate protocols, preventable death and other errors.

@@

Due to lack of data on FAST, type of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[(13)](14)(15).
Due to the tenth edition of ATLS classifications lack actual numbers, we will be approximating numbers to the values. For HR and RR we will be using the numbers from the ninth edition of the classification. The SBP will be divded into class I and II where the SBP is >110, class III 110-90mmhg and class IV <90mmhg. This selection is based on tenth edition, where SBP is defined as normal in class I to II and normal/below in class III and clearly below normal in class IV. These two studies, redefined hypotension as 110mmhg respective 90mmhg dependent on age[(17)](18), therefore we choose class III as 110-90 mmhg and class IV below 90mmhg. We decided to not include Glasgow Coma Scale (GCS) due to difficulties in differentiating if any lowering is due to shock or traumatic brain injury (TBI). Urine output and pulse pressure were also neglected, due to lack of data.

To classify, all of the parameters in one class needs to be fulfilled. In the cases where the parameters belong in different classes, we will be classifying according to the most severe parameter in the same way as Dunham et al(16). At same time ATLS manual states one of the common pitfalls is “Diagnosis of shock can be missed when only a single parameter is used.” therefore to improve sensitivity we decided to use the worst parameter for classification.

We will be only using the base deficit parameter according to the tenth edition of the ATLS classification for haemorrhage. There are three reasons. The first one is ATLS tenth edition lack numerical values for vital parameters. The second one is that bace deficit is superior to vital parameters as a predictive parameter for mortality(16). The third one is due to nature of having multiple parameters, makes it difficult to categorise a patient with vital parameters in two or more classes and difficult to separate lower GCS due to traumatic brain injury from shock as cause. The classes will be divided into class I: 0 - (-2), class II: (-2)-(-6), class III: (-6)-(-10), and class IV: <(-10).

@@

The trauma care quality database is build up in a way where every error can be classified into a certain type of OFI, however for paitent that died there is only the category “preventable death”. Therefore, survivorship bias may occur as we dont have the detail which type of OFI actually occured.

@@

All available data in trauma registry and trauma care quality database will be included.

@@

The statistical analysis will be performed using R, a programming language and environment for statistical computing. We will present the type of OFI as percentage distributions and then visualize it in a pie chart. Unadjusted and adjusted logistic regression will be used to determine the association between the OFI, degree/class of shock according to ATLS and other patient factors such as age, sex, preinjury ASA and ISS. It will be presented as odds ratios (OR) between presence of OFI, and shock classes. The OR will be determined with 95% confidence intervals, and a significance level of 5% will be used. All statistical analysis will first be done on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database to ensure objectivity.

# Ethics

The handling of registry data and the extraction of information from medical records pose a potential risk to patient privacy. To mitigate this, patients were anonymized after data collection, with personal identity numbers replaced by unique study identifiers.

From a self-determination standpoint, the Swedish trauma registry does not require patient consent for data registration. However, it is mandatory for the caregiver to inform patients of their inclusion in the registry. Patients have the right to view, correct, opt out of, and request the deletion of their data, without any impact on the care they receive.

In this study, data will be presented at the group level, ensuring that no individual patient details can be identified. Ethical approvals has been obtained for this study, with the record numbers: 2021-02541, 2021-03531 and 2023-02975-02.

The benefits of this study, in relation to its risks, are substantial due to its potential to improve care for future critically ill patients. This research is a step toward reducing mortality and morbidity in a significant portion of the younger and economically active population worldwide.

@@

After discussion with my supervisor, it seems like the part which have the highest possibility of taking too much time is coding, and therefore he and other cosupervisor will assist with coding if needed.
